Wait-free Queues with Polylogarithmic Step Complexity

### 2022-08-26 thesis/thesis.tex
apply new content structure

\begin{document}

@@

\section{Introduction}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. This is why distributed algorithms have become important. It is not hard to see why multiple processes cannot update sequential data structures designed for one process. For example, consider two processes trying to insert some values into a sequential linked list simultaneously. Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node and after that $q$ does the same. In this run, $p$'s update is overwritten. One solution is to use locks; whenever a process wants to do an update or query on a data structure, the process locks it, and others cannot use it until the lock is released. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the one process holding the lock can make progress.

@@

\section{Related Work}
\subsection{List-based Queues}
In the following paragraphs, we look at previous lock-free queues.
Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue. A version of it is included in the standard Java Concurrency Package. Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}). Head points to the first node in the linked list that has not been dequeued, and Tail points to the last element in the queue. To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}. If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry. This makes the amortized number of steps to be $\Omega(p)$ per enqueue. Similarly, dequeue can take $\Omega(p)$ steps.

@@

\pagebreak
\section{ Universal Construction using Tournament Tree with Big CAS Objects }

\paragraph{}
A universal construction gets any sequential object and creates a concurrent version of the given datastructure. Our universal construction in Algorithm~\ref{alg1} relies on a tournament tree with $p$ leaves and height $\log(p)$ is shared among $p$ processes. Nodes in the tree are CAS objects that store an ordering and support \tt{append()} and \tt{diff()} operations. Leaf $l_i$ stores the sequence of the operations invoked by $P_i$. Each internal node stores the sequence of operations propagated up to it. When process $P_i$ wishes to apply an operation $op$ on the implemented object, it appends $op$ to its assigned leaf and tries to propagate it up to the root. The history of operations stored at the root is the linearization ordering. The operation $op$ is linearized when it is appended to the root.

\begin{center}
\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
\end{center}

The algorithm uses a subroutine \textsc{Refresh}($n$) that concatenates new operations from node $n$'s children (that have not already been propagated to $n$) to the sequence of operations stored in $n$ and tries to CAS the new sequence into $n$. In other words, \textsc{Refresh}($n$) tries to append $n$'s children's new operations to $n$'s sequence.
After a process adding a new operations to its leaf, it has to propagate new operations up to the root. \textsc{Propagate}($n$) tries to append $n$'s new operations to the root $n$ by recursively calling \textsc{Refresh}($n$). In each \tt{Propagate()} step if a \textsc{Refresh}($n$) fails, it means another CAS operation has succeeded; if so, it tries to \textsc{Refresh}($n$) again. If the second attempt fails too, another process has already appended the operations the current \textsc{Propagate} is trying to append.
Operations that were in $l_i$ before \textsc{Propagate}($l_i$.parent) was invoked are guaranteed to be added to the root by the time the \textsc{Propagate}($l_i$.parent) terminates.

\begin{figure}
\begin{center}
\begin{subfigure}[b]{.49\textwidth}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
\Tree [.{\begin{tabular}{|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ \\ \hline \end{tabular}}
{\begin{tabular}{|c|c|c||c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
{\begin{tabular}{|c|c||c|c|}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
\end{tikzpicture}
}
  \caption{Operations after $||$ are new.}

\end{subfigure}
\hfill
\begin{subfigure}[b]{.49\textwidth}

  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tikzpicture}[level 1/.style={level distance=1.4cm,sibling distance=0.5cm}]
\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{r_1}$ & $op_{l_3}$ & $op_{r_2}$ & $op_{l_4}$ & $op_{l_5}$ & $op_{r_3}$ & $op_{r_4}$  \\ \hline \end{tabular}}
{\begin{tabular}{|c|c|c|c|c||}  \hline $op_{l_1}$ & $op_{l_2}$ & $op_{l_3}$ & $op_{l_4}$ & $op_{l_5}$ \\ \hline \end{tabular}}
{\begin{tabular}{|c|c|c|c||}  \hline $op_{r_1}$ & $op_{r_2}$ & $op_{r_3}$ & $op_{r_4}$\\ \hline \end{tabular}} ]
\end{tikzpicture}
}
  \caption{New operations are added to the parent node.}

\end{subfigure}

\caption{\label{fig:ucexample} Propagate Step in Universal Construction}
\end{center}
\end{figure}

\begin{algorithm}
\caption{Universal Construction Idea}\label{alg1}
\begin{algorithmic}[1]
\begin{multicols}{2}

\Function{response}{Do}{\textsl{operation} op, \textsl{pid} i} 
\State \texttt{l\textsubscript{i}.\Call{append}{op}}
\State \Call{Propagate}{parent of l\textsubscript{i}}
\State \texttt{Run} \textsf{the sequence stored in root}
\State \texttt{\Return op}\textsf{'s response from line 4}
\EndFunction{Do}

\Function{void}{Propagate}{\textsl{node} n}
\If{\texttt{n==root}} \Return
\ElsIf{!\Call{Refresh}{n}}
\State \Call{Refresh}{$n$} \EndIf
\State \Call{Propagate}{parent of n}
\EndFunction{Propagate}

\columnbreak

\Function{boolean}{Refresh}{\textsl{node} n}
\State \texttt{old=} \Call{Read}{n}
\State \texttt{new=} \textsf{ops that n's children contain but \texttt{old} does not}
\State \texttt{new= old$\cdot$new}
\State \Return n.\Call{CAS}{old, new}
\EndFunction{Refresh}

$\textsc{O}(\log n)$ CAS operations are invoked to do a \textsc{Propagate}, but the CAS words store sequences of unbounded length.
The problem is that we are trying to store unbounded sequence of operations in each node $n$ (see Figure~\ref{fig::uc}). However, to compute the result of an operation, we only use the total ordering that is stored at the root. Although we use a similar construction for our queue implementation, we develop an implicit representation of the sequence of operations, so that we can use reasonable sized CAS objects and still achieve polygarithmic step complexity.

\begin{figure}[h]
\begin{center}
\begin{tikzpicture}[level 1/.style={level distance=1.8cm,sibling distance=1cm}, level 2/.style={level distance=1.4cm,sibling distance=1cm}]

\Tree [.{\begin{tabular}{|c|c|c|c|c|c|c|c|c|c}  \hline  $op_2^1$ & $op_2^2$ & $op_3^1$ & $op_4^1$ & $op_3^2$ & $op_1^1$ & $op_4^2$ & $op_1^2$ \\ \hline\end{tabular}}  [.{\begin{tabular}{|c|c|c|c|c|c}  \hline $op_2^1$ & $op_2^2$ & $op_1^1$ & $op_1^2$ \\ \hline\end{tabular}}
      {\begin{tabular}{|c|c|c|c}  \hline $op_1^1$ & $op_1^2$ \\ \hline\end{tabular}} {\begin{tabular}{|c|c|c|c}  \hline $op_2^1$ & $op_2^2$ \\ \hline\end{tabular}} ] [.{\begin{tabular}{|c|c|c|c|c|c}  \hline $op_3^1$ & $op_4^1$ & $op_3^2$ & $op_4^2$ \\ \hline\end{tabular}} {\begin{tabular}{|c|c|c|c}  \hline $op_3^1$ & $op_3^2$ \\ \hline\end{tabular}} {\begin{tabular}{|c|c|c|c}  \hline $op_4^1$ & $op_4^2$ \\ \hline\end{tabular}} ] ]

\end{tikzpicture}
\caption{\label{fig::uc} Universal Construction: $op_j^i$ denotes the $i$th operation from process $j$. In each node, we store the ordering of all the operations propagated up to it.}
\end{center}
\end{figure}

\pagebreak
\section{Block Trees}
Here we introduce a data structure that allows processes to agree on the linearization ordering of their operations using $\textsc{O}(\log p)$ \texttt{CAS} per operation called a \textit{block tree}. Then we use the block tree as a stepping stone towards our queue algorithm.
A block tree is a tournament tree shared among $p$ processes (see Figure~\ref{fig::blocktree}). Each process has a leaf, and it appends its operations to its leaf. After that, the process tries to propagate its new operation up to the tree's root. An ordering of operations propagated up to a node is stored in that node. All processes agree on the sequence stored in the root and this is used as the linearization ordering. Our idea is similar to Jayanti and Petrovic's multi-enqueuer single-dequeuer Queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use \texttt{CAS} operations with big words and do not put a limit on the number of concurrent operations.

@@

%\subsection{Idea in a nutshell}
We apply two ideas from universal construction to create a new linearizable data structure agreeing on a sequence of elements among processes. First, there is a shared tournament tree among processes, in which each process appends its element to its leaf in the tree and then tries to propagate it up to the root by performing \textsc{Refresh()} operations at each node. Second, each operation is linearized when its element is appended to the root.

\paragraph{}
 In the universal construction, we order new concurrent operations at each \textsc{Refresh}() and maintain that order in the path up to the root. However, we can instead keep track of sets of concurrent operations and create the total ordering of all operations at the root (see Figure~\ref{fig::set}).

@@

%\begin{lemma}
%Operations in the same block can be linearized in any order.
%\end{lemma}

\paragraph{}
 We used unbounded CAS objects storing sequences as big words in the universal construction. One can represent sequences as arrays to overcome this implementation problem. Each array element will store one of the blocks of concurrent operations described in section \ref{subsec::block}.

\paragraph{}
Copying operation sequences from children to their parent in a \textsc{Refresh}() takes time proportioned to the number of operations being copied. This is time-consuming, so we propose a way to augment the tree to calculate lines 15,16 in $\textsc{O}(\log p)$ steps which reads new operation and concats them with old operations. Instead of representing the set of operations by explicitly listing them in a node, we represent a set of operations implicitly by recording which of the children's sets were unioned to create the set. Having operation sequences stored at leaves, we can deduce a set of operations in a node using this implicit representation.~(see Figure~\ref{fig:block}.)

\pagebreak
\begin{figure}

@@

Each block $b$ in node $n$ is the aggregation of blocks in the  children of $n$ that are newly read by the\textsc{Propagate}() step that created block $b$. For example, the third block in the root (8,36) is created by merging block (5,3) from the left child and (14,6) and (0,16) from the right child. Block (5,3) also points to elements from blocks (0,5) and (1,2). 
%Recursively we can deduce each block contents until the leaf nodes store arrays of individual element operations.

\begin{definition}\{Existence of an operation in a block\}  Operation op exists in block b if it has propagated up to block b.
\end{definition}

\begin{definition}\{Subblock\}
  The blocks that are aggregated into block $b$ in a \textsc{Propagate}() step are called subblocks of $b$. Block $b_1$ is a subblock of $b_2$ if and only if $b_1$ is a block in node $v$ and in $b_2$ is a block in the parent of $v$ and $b_1$'s elements exits in $b_2$'s elements.
\end{definition}
%{\TODO: describe how blocks work with a good example how exactly it is done}

\paragraph{}
We choose to linearize operations in a block from the left child before those from the right child as a convention. Operations within a block of the root can be ordered in any way that is convenient. In effect, this means that if there are concurrent new blocks in a \textsc{Refresh}() step from several processes we linearize them in the order of their process ids. So for example  operations aggregated in block (10,2) are in the order (2,3),(4,1),(0,2). All blocks from the left child with come before the right child and the order of blocks of each child is preserved among themselves.
%\begin{lemma}
%\label{lem:block_size}
%There cannot be more than one operation from each process in a \textsc{Refresh}() invocation.
%\end{lemma}
%\begin{proof}
%  If so, then that parent process has invoked two simultaneous operations. $ \blacksquare $
%\end{proof}

\paragraph{}
In a \textsc{Propagate}() invocation path from a leaf to root, there will be \textsc{Refresh}() steps with merges from $2, 4, 8, ..., p$ processes. So in a complete propagation, at most $2p$ blocks are merged into one block. \texttt{(maybe useful for analysis)}
%\begin{lemma}
%A block in the root cannot contain more than one operation from a process.
%\end{lemma}
%\begin{proof}
%  Assume a process in a block invokes two operations; one has to be finished another. If so, it has been added to root before. $\blacksquare$
%\end{proof}

\pagebreak
\subsection{Using pointers and prefix sum to make \textsc{GetIndex}($i$) faster}
%\begin{definition}
%  \textsc{BSearch}($i$): Search $i$th element in a prefix sum array with length $n$ in \textsc{O}($\log n$) steps.
%\end{definition}
\paragraph{}
\textsc{GetIndex}($i$) returns the $i$th operation stored in the block tree sequence. We do that by finding the block $b_i$ containing $i$th element in the root, and then recursively finding the subblock of $b_i$ which contains $i$th element. To make this recursive search faster, instead of iterating over all elements in sequence of blocks we store prefix sum of number of elements in the blocks sequence and pointers to make BSearch faster.

@@

\pagebreak
\subsection{Block Tree Algorithm}
\paragraph{}
Our Block Tree is a linearizable implementation of a data structure that stores a sequence of elements. It has two methods (see Algorithm~\ref{alg2}), \textsc{Append}($e$) which appends element $e$ to the sequence, and \textsc{Get}($i$) which returns the $i$th element in the sequence.

\paragraph{Design of a block tree}

@@

\paragraph{Specification}
A block tree is a shared data structure that stores a sequence of elements.  It has two methods \texttt{Append(e)} and \texttt{Get(i)}. \texttt{Append(e)} adds \texttt{e} to the end of the sequence and returns the index of \texttt{e} in the sequence. \texttt{Get(i)} returns \texttt{i}th element stored in the sequence.

\paragraph{SubBlock} Block \texttt{s} is a subblock of \texttt{b} if \texttt{s} is between blocks \texttt{start..end} in \texttt{n} from Lines 41,42 of \texttt{CreateBlock()}.

\paragraph{Membership}
Element \texttt{e} is a member of block \texttt{b} in:
\begin{itemize}
  \item internal node \texttt{n}, if \texttt{e} is a memeber of \texttt{s} that \texttt{s} is a subblock of \texttt{b}.
  \item leaf node \texttt{n}, if \texttt{e} belongs to \texttt{n.dir.blocks[b\textsuperscript{$\prime$}.end\textsubscript{dir}+1..b.end\textsubscript{dir}]} for \texttt{dir} $\in$ \texttt{\{left, right\}} which b\textsuperscript{$\prime$} is the previous block of \texttt{b} in \texttt{n}.
\end{itemize}

\pagebreak
\paragraph{Order of elements inside node}
Element \texttt{d} is before element \texttt{e} in node \texttt{n}, if:
\begin{itemize}
  \item The block containing d is before the block containing e.
  \item e and d are in the same block and d is in the left child and e is in the right child.
  \item d is before e in the same child's order.
\end{itemize}

@@

\pagebreak
\paragraph{Double Refresh}
Elements in \texttt{n}'s children's blocks before line \texttt{13} are guaranteed to be in \texttt{n}'s blocks after Line~\texttt{15}.
\begin{proof}
\texttt{CreateBlock()} reads blocks in the children that does not exist in the parent and aggregates them into one block. If a \texttt{Refresh} procedure returns true it means it has appended the block created by \texttt{CreateBlock()} into the parent node's sequence. So suppose two \texttt{Refresh}es fail. Since the first \texttt{Refresh} was not successful, it means another CAS operation by a \texttt{Refresh}, concurrent to the first\texttt{Refresh}, was successful before the second \texttt{Refresh}. So it means the second failed \texttt{Refresh} is concurrent with a successful \texttt{Refresh} that assuredly has read block before the mentioned line \texttt{13}. After all it means if any of the \texttt{Refresh} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.
\end{proof}
\begin{figure}[hbt]
  \center\includegraphics[width=4in]{pics/doublerefresh.png}
  \caption{The second failed Refresh is assuredly concurrent to a Successful \texttt{Refresh()} with \texttt{CreateBlock} line after first failed \texttt{Refresh}'s \texttt{CreateBlock()}.}
\end{figure}

\paragraph{Disjunction} Blocks in node \texttt{n}'s contain disjoint sets of elements.
\begin{proof}
Without loss of generality, assume blocks \texttt{b1, b2} contain common element \texttt{e} from the left child, and \texttt{b2} is after \texttt{b1} in \texttt{n}'s sequence of blocks. So block start of b2's \texttt{CreateBlock()} is after block end of \texttt{b1}'s end. Since \texttt{b2}'s start is the end of the block before itself, it cannot be before \texttt{b1}'s end.
\end{proof}

\paragraph{Total Order}
Sequence represented by the Block Tree is the sequence of the blocks stored in the root.

\paragraph{Linearization Points}
\texttt{Get(i)} is linearized when it terminates. \texttt{Append(e)} is linearized right after when a block containing \texttt{e} is appended to the root, if there are multiple elements appended together, they are linearized by the defined order in the root.

\paragraph{Subblocks Upperbound} Block b has at most $p$ subblocks.
\begin{proof}
If there are more than $p$ subblocks, then there is more than one block from process pl. \texttt{Append(e)} finishes after propagating and appending \texttt{e} to the root(line 9). So these blocks cannot be appended to root already, so pl has invoked two concurrent \texttt{Append()}s(line 1) without terminating the first one.
\end{proof}

@@

\paragraph{Computing superblock}
\begin{enumerate}
  \item Value read for \texttt{super[b.time]} in line 71 is not null.\begin{proof}
    \texttt{Index()} is invoked after finishing \texttt{Propagate()} in line 10. For each value \texttt{c\textsubscript{dir}} read in lines 23, \texttt{super} is set before incrementing in lines 26,27.
  \end{proof} 
  \item \texttt{super[]} preserves order from child to parent; if in a child block \texttt{b} is before \texttt{c} then \texttt{b.time} $\leq$ \texttt{c.time} and \texttt{super[b.time]} $\leq$ \texttt{super[c.time]}\begin{proof}
   Follows from the order of lines 37, 26, 27.
 \end{proof}
%  \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
%  \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
  \item \texttt{super[i+1]-super[i]}$\leq p$
\begin{proof}
 In a Refresh with successful CAS in line 24, \texttt{super} and \texttt{counter} are set for each child in lines 26,27. Assume the current value of the counter in node \texttt{n} is \texttt{i+1} and still \texttt{super[i+1]} is not set. If an instance of successful \texttt{Refresh(n)} finishes \texttt{super[i+1]} is set a new value and a block is added after \texttt{n.parent[sup[i]]}. There could be at most $p$ successful unfinished concurrent instances of \texttt{Refresh()} that have not reached line 27. So the distance between \texttt{super[i+1]} and \texttt{super[i]} is less than $p$.
 \end{proof}
  \item Superblock of \texttt{b} is within range $\pm 2p$ of the \texttt{super[b.time]}.
\begin{proof}
\texttt{super[i]} is the index of the superblock of a block containing block b. It is trivial to see that \texttt{n.super} and \texttt{n.b.counter} are increasing.  \texttt{super(b)} is the real superblock of b. \texttt{super(t]} is the index of the superblock of the last block with time \texttt{t}. If \texttt{b.time} is \texttt{t} we have:
$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
  \end{proof} 
\end{enumerate}

\section{Implementing Queue using Block Tree}
In this work, we design a queue with $O(\log^2 p +\log n)$ steps per operation, where $n$ is the number of total operations invoked. We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays (CAS Retry Problem). A queue stores a sequence of elements and supports two operations, enqueue and dequeue. \texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored.\texttt{Dequeue()} removes and returns the first element among in the sequence. If the queue is empty it returns \texttt{null}. Knowing index $i$ is the tail of the queue, we can return the dequeue response using \texttt{Get(i)}.  So in the rest we modify block tree to compute \texttt{i} for each \texttt{Dequeue()} to achieve a FIFO queue.
%
%\subsection{What should a \texttt{Dequeue()} return, regarding history of operations?}
\paragraph{}
Next, we describe how to use block tree to implement queues. The block tree, maintains the history of all operations, not only the current state of the queue. Now consider the following history of operations. What should each \texttt{Dequeue()} return? We can implement Enqueue and Dequeue using our block tree. An \texttt{Enqueue(e)} appends an operation with input argument \texttt{e} in the block tree. To do a \texttt{Dequeue()}, process $p$ first appends a \texttt{DEQ} operation to the tree. Then $p$ finds the rank of the \texttt{DEQ} using \texttt{Index()}, the rank of the \texttt{DEQ} and the information stored in the root about the queue $p$ computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}. Finally $p$ returns the argument of that \texttt{ENQ} using \texttt{Get(i)}.

@@

\begin{definition}
A non-null dequeue is one that returns a non-null value.
\end{definition}

\paragraph{}
In the example above, \texttt{Dequeue()} operations return \texttt{5, 2, 3, null, 4} in order. Before \texttt{ENQ(4)} the queue gets empty so the last \texttt{DEQ()} returns null. If the queue is non-empty and $r$ \texttt{Dequeue()} operations have returned a non-null response, then $i$th \texttt{Dequeue()} returns the input of the $r+1$th \texttt{Enqueue()}. So, in order to answer a Dequeue, it's sufficent to know the size of the queue and the number of previous non-null dequeues.

@@

\pagebreak
\section{Main Algorithm}

@@

\paragraph{Tournament Tree}
In order to reach an agreement on the order of operations among $p$ processes, we use a Tournament Tree. Leaf \texttt{l\textsubscript{i}} is assigned to a process \texttt{i}. Each process adds $op$ to its leaf. In each internal node an ordering of operations in its subtree is stored. All processes agree on the total ordering of all operations stored in the root. This ordering will be the linearization of the operations.
\paragraph{Implicit Storing Blocks}
For efficiency, instead of storing explicit sequence of operations in nodes of the Tournament Tree, we use Blocks. A Block is a constant size object that implicitly represents a sequence of operations. In each node there is an array of Blocks.

\begin{definition}[Block]
A block is an object that stores some statistics described in Algorithm Queue.   
\end{definition}

\begin{definition}[\label{def::subblock}Subblock]
Block \texttt{b} is a subblock of \texttt{n.blocks[i]} if it is in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} or \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.
\end{definition}
Block $b$ contains subblocks in the left and right children. WLOG left subblocks of $b$ are some consecutive blocks in the left child starting from where previous block of $b$ has ended to the the end of $b$. See Figure \ref{fig::createBlock} .

\begin{definition}[Membership of an operation in a block]
Operation \texttt{e} is a member of block \texttt{b} in:
\begin{itemize}
 \item leaf node \texttt{n}, if \texttt{e} belongs to \texttt{n.ops[b's index]}.
 \item internal node \texttt{n}, if \texttt{e} is a member of \texttt{s} that \texttt{s} is a subblock of \texttt{b}.
\end{itemize}

\end{definition}

@@

\pagebreak
\section{Pseudocode}

@@

\section{Proof of Linearizability}

@@

\begin{lemma}[CASes invoked]
An \nf{Enqueue()} or \nf{Dequeue()} operation, does at most $4\log p$ \nf{CAS} operations.
\end{lemma}
\begin{proof}
  In each height of the tree at most 2 times \nf{Refresh()} is invoked and  every \nf{Refresh()} has 2 \nf{CAS}es, one in Line \ref{cas} and one in Lines \ref{incrementHead1} or \ref{incrementHead2}.
\end{proof}
\end{document}

>>>>

\begin{document}
\title{Wait-free Queues with Polylogarithmic Step Complexity}
\author{Hossein Naderibeni \\supervised by Eric Ruppert}
\maketitle

@@

\tableofcontents
\pagebreak
\section{Introduction}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. This is why distributed algorithms have become important. It is not hard to see why multiple processes cannot update sequential data structures designed for one process. For example, consider two processes trying to insert some values into a sequential linked list simultaneously. Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node and after that $q$ does the same. In this run, $p$'s update is overwritten. One solution is to use locks; whenever a process wants to do an update or query on a data structure, the process locks it, and others cannot use it until the lock is released. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the one process holding the lock can make progress.

@@

\pagebreak
\section{Related Work}
\subsection{List-based Queues}
In the following paragraphs, we look at previous lock-free queues.
Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue. A version of it is included in the standard Java Concurrency Package. Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}). Head points to the first node in the linked list that has not been dequeued, and Tail points to the last element in the queue. To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}. If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry. This makes the amortized number of steps to be $\Omega(p)$ per enqueue. Similarly, dequeue can take $\Omega(p)$ steps.

@@

\subsection{Attiya Fourier Lower Bound}
\pagebreak

\pagebreak
\section{Our work}
Jayanti introduced a wait-free polylogarithmic MESD queue, we are going to design a polylogarithmic MEMD queue based on some of their ideas. We apply two ideas from universal their work to create a new linearizable data structure agreeing on a sequence of elements among processes. First, there is a shared tournament tree among processes, in which each process appends its element to its leaf in the tree and then tries to propagate it up to the root by performing \textsc{Refresh()} operations at each node. Second, each operation is linearized when its element is appended to the root.  We use the shared tournament tree among $p$ processes (see Figure~\ref{fig::blocktree}) to agree on one total ordering on the operations invoked by procesess. Each process has a leaf, and it appends its operations to its leaf. After that, the process tries to propagate its new operation up to the tree's root. An ordering of operations propagated up to a node is stored in that node. All processes agree on the sequence stored in the root and this is used as the linearization ordering. Our idea is similar to Jayanti and Petrovic's multi-enqueuer single-dequeuer Queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use \texttt{CAS} operations with big words and do not put a limit on the number of concurrent operations.

@@

\paragraph{}
We can instead keep track of sets of concurrent operations and create the total ordering of all operations at the root (see Figure~\ref{fig::set}).

@@

\begin{figure}

@@

In the original algorithm  we differ between the enqueues and the dequeues in a block. Later we discuss about the reason.

Each block $b$ in node $n$ is the aggregation of blocks in the  children of $n$ that are newly read by the\textsc{Propagate}() step that created block $b$. For example, the third block in the root (8,36) is created by merging block (5,3) from the left child and (14,6) and (0,16) from the right child. Block (5,3) also points to elements from blocks (0,5) and (1,2).

\paragraph{}
We choose to linearize operations in a block from the left child before those from the right child as a convention. Operations within a block of the root can be ordered in any way that is convenient. In effect, this means that if there are concurrent new blocks in a \textsc{Refresh}() step from several processes we linearize them in the order of their process ids. So for example  operations aggregated in block (10,2) are in the order (2,3),(4,1),(0,2). All blocks from the left child with come before the right child and the order of blocks of each child is preserved among themselves.

\paragraph{}
In a \textsc{Propagate}() invocation path from a leaf to root, there will be \textsc{Refresh}() steps with merges from $2, 4, 8, ..., p$ processes. So in a complete propagation, at most $2p$ blocks are merged into one block. \texttt{(maybe useful for analysis)}

\paragraph{}
\textsc{GetIndex}($i$) returns the $i$th operation stored in the block tree sequence. We do that by finding the block $b_i$ containing $i$th element in the root, and then recursively finding the subblock of $b_i$ which contains $i$th element. To make this recursive search faster, instead of iterating over all elements in sequence of blocks we store prefix sum of number of elements in the blocks sequence and pointers to make BSearch faster.

@@

\pagebreak
\subsection{Queue Decription}
\paragraph{Design of a Queue}

@@

@@

\begin{figure}[hbt]
  \center\includegraphics[width=4in]{pics/doublerefresh.png}
  \caption{The second failed Refresh is assuredly concurrent to a Successful \texttt{Refresh()} with \texttt{CreateBlock} line after first failed \texttt{Refresh}'s \texttt{CreateBlock()}.}
\end{figure}

@@

\paragraph{Computing superblock}

\paragraph{Implementing Queue using Block Tree}
In this work, we design a queue with $O(\log^2 p +\log n)$ steps per operation, where $n$ is the number of total operations invoked. We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays (CAS Retry Problem). A queue stores a sequence of elements and supports two operations, enqueue and dequeue. \texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored.\texttt{Dequeue()} removes and returns the first element among in the sequence. If the queue is empty it returns \texttt{null}. Knowing index $i$ is the tail of the queue, we can return the dequeue response using \texttt{Get(i)}.  So in the rest we modify block tree to compute \texttt{i} for each \texttt{Dequeue()} to achieve a FIFO queue.
%
%\subsection{What should a \texttt{Dequeue()} return, regarding history of operations?}
\paragraph{}
Next, we describe how to use block tree to implement queues. The block tree, maintains the history of all operations, not only the current state of the queue. Now consider the following history of operations. What should each \texttt{Dequeue()} return? We can implement Enqueue and Dequeue using our block tree. An \texttt{Enqueue(e)} appends an operation with input argument \texttt{e} in the block tree. To do a \texttt{Dequeue()}, process $p$ first appends a \texttt{DEQ} operation to the tree. Then $p$ finds the rank of the \texttt{DEQ} using \texttt{Index()}, the rank of the \texttt{DEQ} and the information stored in the root about the queue $p$ computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}. Finally $p$ returns the argument of that \texttt{ENQ} using \texttt{Get(i)}.

@@

\paragraph{}
A non-null dequeue is one that returns a non-null value. In the example above, \texttt{Dequeue()} operations return \texttt{5, 2, 3, null, 4} in order. Before \texttt{ENQ(4)} the queue gets empty so the last \texttt{DEQ()} returns null. If the queue is non-empty and $r$ \texttt{Dequeue()} operations have returned a non-null response, then $i$th \texttt{Dequeue()} returns the input of the $r+1$th \texttt{Enqueue()}. So, in order to answer a Dequeue, it's sufficent to know the size of the queue and the number of previous non-null dequeues.

@@

@@

\paragraph{Tree}
In order to reach an agreement on the order of operations among $p$ processes, we use a Tournament Tree. Leaf \texttt{l\textsubscript{i}} is assigned to a process \texttt{i}. Each process adds $op$ to its leaf. In each internal node an ordering of operations in its subtree is stored. All processes agree on the total ordering of all operations stored in the root. This ordering will be the linearization of the operations.
\paragraph{Implicit Storing Blocks}
For efficiency, instead of storing explicit sequence of operations in nodes of the Tournament Tree, we use Blocks. A Block is a constant size object that implicitly represents a sequence of operations. In each node there is an array of Blocks.

Block $b$ contains subblocks in the left and right children. WLOG left subblocks of $b$ are some consecutive blocks in the left child starting from where previous block of $b$ has ended to the the end of $b$. See Figure \ref{fig::createBlock} .

@@

\pagebreak
\subsection{Pseudocode}

@@

\subsection{Proof of Correctness}

@@

\begin{lemma}[CASes invoked]
An \nf{Enqueue()} or \nf{Dequeue()} operation, does at most $4\log p$ \nf{CAS} operations.
\end{lemma}
\begin{proof}
  In each height of the tree at most 2 times \nf{Refresh()} is invoked and  every \nf{Refresh()} has 2 \nf{CAS}es, one in Line \ref{cas} and one in Lines \ref{incrementHead1} or \ref{incrementHead2}.
\end{proof}

\pagebreak
\subsection{Garbage Collection or Getting rid of the infinite Arrays}

\pagebreak
\section{Using Queues to Implement Vectors}
Supporting Append, Read, Write in PolyLog time by modifying Get(Enq) Method. Create a Universal Construction Using our vector

\pagebreak
\section{Conclusion}
possible directions for work

Maybe Stacks

Characterize what datastructure can be used for this approach, we already know: queue, fetch \& Inc, Vectors

\end{document}
